Refinements for Open Automata

### 2023-06-26 2021-Refinement/paper/refinementarticleV2.tex
figs
improvements

\keywords{Labelled transition systems   \and Simulation \and Composition.}
\end{abstract}
%
%
%
\section{Introduction}
%
%
%\TODO{TODO list}
%
%
%* \st{example a terminer}
%
%* choisir terminologie: refinement or simulation
%
%* \st{open automata  / Open Automata -> only use open automata or OA.}
%
%* positioning
%
%* \st{format according to conf: lncs FACS or SEFM}
%
%* global rewriting (partly done)

@@

%\begin{definition}[Expression algebra, Action algebra, Formulas, Terms]
Our expression algebra \(E\) is the disjoint union  of  terms,  actions, and  formulas
\( E=\terms \uplus \actions \uplus \formulas\).
\(\terms\) and \(\actions\) are term algebras.
The formulas \(\formulas\) contain at least first order formulas and equality\footnote{Equality does not need to be only syntactic.} over \(\terms\) and \(\actions\). 
%\end{definition}
For \(e \in E\),
 \(\fvars{e}\) is the set of variables in  $e$ that are not bound by a binder. 
An expression is closed if \(\fvars{e}=\emptyset\).
The set \(\values\) denotes values which is a subset of closed terms. \(\rformulas[V]\) is the set of formulas $f$ that only uses variables in $V$, i.e. the formulas such that  \(\fvars{f}\subseteq V\).
%The substitution in \(e \in E\) of \(x \in \fvars{e}\) by \(t \in \terms\), is denoted \(e\subst{t}{x}\), and its generalisation to t
The parallel substitution of variables in \(V\) by a map \(\psi: V \to \terms\) is denoted \(e\psubst{\psi}\).

@@

@@

The OA  modelling the behaviour of such a system  using an  unbounded circular/ring buffer is depicted in Fig. \ref{Fig:RefineOA}.  The automaton has a single state with  two holes: $\texttt{P}$ and  $\texttt{Q}$ that are the two interacting processes.  $\texttt{l}$ (as last)  indicates the next available position for  enqueuing an element  and  $\texttt{f}$ (as first) is the position that contains the next element to be dequeued. %\footnote{The  arrow  without any state of origin is a graphical convention to indicate that the pointers are initialized at the beginning}. 
The buffer reacts to a push from $\texttt{P}$ and enqueues it. Similarly, whenever   $\texttt{Q}$ pops an element, it dequeues it. Additionally, whenever $\texttt{Q}$ produces an item, it is exposed as an external observable  $\texttt{print}$ action. 
When any process do its internal computation, it is exposed externally as unobservable action $\tau$. 
\begin{figure}[!tb]
 \centering
   \includegraphics[width=.99\textwidth]{Figures/FIFORing.pdf}
   \caption{OA for the \texttt{prod-cons} system using FIFO circular buffer.
    \label{Fig:RefineOA}} 
\end{figure}
\qed
\end{example}

@@

\begin{example} \label{Example:prodandcons} Fig. \ref{Fig:procandcons} shows a producer automaton  and a consumer  automaton that can be used  to fill the holes $\texttt{P}$ and $\texttt{Q}$ of \texttt{prod-cons} defined in Example~\ref{Example:fifo-system}.
 \begin{figure}[!b]
 \centering
   \includegraphics[width=.35\textwidth]{Figures/P-proc.pdf}\hfill 
   \includegraphics[width=.45\textwidth]{Figures/Q-proc.pdf}
   \caption{(Left) A \texttt{producer}.  It produces one item at a time and pushes it.  (Right)  A \texttt{consumer}. It pops an item, does some work and pushes the result. \label{Fig:procandcons}}
\end{figure}

@@

\begin{figure}[!b]
 \centering
   \includegraphics[width=.99\textwidth]{Figures/compos-FIFO-Pro.pdf}
   \caption{OA for \texttt{prod-cons}[\texttt{P}/\texttt{producer}]. 
   \label{Fig:ComposeOA}} 
\end{figure}
\end{example}

\subsection{Relations between Open Automata}
Establishing semantic equivalences and simulation relations between different OA requires  to compare their states. For this purpose, we suppose that the variables of the two OA are disjoint (a renaming of variables may have to be applied before comparing OA states).
\begin{definition}[Relation on open automata configurations] Suppose $V_1$ and $V_2$ are disjoint.
A relation on configurations of two OA \(\OAg[1]\) and \(\OAg[2]\) is a function \( \mathcal{R}: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\).
\end{definition}
%\TODO{C'est etrange, la formulation "the variables they refer to", comme il n'y a plus de variables d'etat... +1}
The idea is that two states are related depending on the satisfiability of the expression relying their variables i.e., if the variables of the OA verify a certain formula. 
In other words, to each pair of states is attached a boolean formula that may refer to the  variables of each of the two OA, stating whether the two states are related or not.
Additionally, we say that the relation $ \mathcal{R}$ \emph{relates the initial states} of the automata if: \(\sigma_{01} \uplus \sigma_{02} \vdash  \mathcal{R}\mpar{s_{01}, s_{02}}\).

@@

Reachability takes into account all paths, and can over-approximate the reachable configurations. 
From an automation point of view, finding the most precise reachability predicate for a given automata is not decidable because of the symbolic nature of OA, but only an over-approximation is necessary.

@@

Similarly to FH-bisimulation~\cite{AMHEEMA:2023} we are interested  in finding simulation relations between configurations of two OA that contain variables and holes. When dealing with open systems it is common to define simulation in terms of a simulation relation.
We rely on a classical notion of simulation inspired from~\cite{AMHEEMA:2023}. The idea is to consider two configurations related by a relation; if one state can do a transition, then the other can do a transition too. 
Like for bisimulation, a simulation relation characterises when two states are related, and this  characterisation is expressed as a predicate on the variables of the two automata.
Simulation defines conditions on a predicate $\mathcal{R}$ such that $\mathcal{R}\mpar{s_1, s_2}$ is a predicate (possibly involving variables of the automata) that is true when the state $s_1$ of $A_1$ refines the state $s_2$ of $A_2$.

@@

\begin{example} \label{example:fifo-system-one}
To illustrate the simulation simulation of OA, we consider a variation on the \texttt{prod-cons} example. Namely,  we suppose that  the two processes  \texttt{P} and  \texttt{Q} communicate through a one-place  buffer. Fig. \ref{Fig:SpecOA} shows the OA modelling this simpler version of the system, that we refer to as \texttt{simprod-cons}. 
\begin{figure}[!b]
 \centering
   \includegraphics[width=.90\textwidth]{Figures/FIFOpen2.pdf}
   \caption{  The \texttt{simprod-cons} OA: the system using one-place  buffer.  \label{Fig:SpecOA}}
\end{figure}
We can easily check that this automaton simulates the one of Fig.~\ref{Fig:RefineOA}.
Indeed, one can see that  $\mathcal{R}=\{(r_0,s_0) \mapsto \texttt{l}=\texttt{f}, (r_0,s_1) \mapsto \texttt{f}=\texttt{l}+1\% \texttt{N} \}$ is  a candidate simulation 
relation. It follows that  \texttt{prod-cons} $\leq$ \texttt{simprod-cons}. \qed
\end{example}

@@

\medskip
One major challenge in the extension of simulation to different sets of holes is to maintain a form of transitivity while being able to take into account the actions of some of the holes. A naive definition of simulation would ensure that only the holes that are identical in the two OA are taken into account in the simulation. Unfortunately considering all the common holes does not ensure transitivity of the simulation for the following reason. If $A_1$ simulates $A_2$ and $A_2$ simulates $A_3$, and one hole $j$ appears in $A_3$ and in $A_1$ but not in $A_2$ then we have no guarantee on the way $A_1$ and $A_3$ take the actions of this his hole into account, thus  a simulation between and $A_1$ and $A_3$ would require conditions involving actions of the hole $j$ which cannot be ensured. The way we solve this issue is to remember in the simulation relation which holes have been compared. This makes the relation parameterised by a subset of the set of holes that belong to the two automata that we want to take into account.
This way, in the example above, we would have no guarantee on actions the hole $j$ by transitivity but can state a simulation relation with guarantees on the actions of the other holes.

@@

\begin{property}[Relating simulations]\label{lem-sim-sim}
Hole-equal simulation is a particular case of hole-tracking simulation when  $J_1=J_2=H$.
\end{property}
\begin{example} Consider the automata of examples \ref{Example:fifo-system} and \ref{example:fifo-system-one}. As we saw above, \texttt{prod-cons} $\leq$ \texttt{simprod-cons}, therefore \texttt{prod-cons} $\leq_{\{\texttt{P,Q}\}}$ \texttt{simprod-cons}.
\end{example}
%\TODO{potential ex 4: check if the example with a hole filled is a simulation of the example with holes? }

@@

Before reasoning on the properties of simulation. we need to introduce one additional notion that characterizes when the composition of two automata does not introduce new blocked transitions.

@@

%\begin{definition}[Non-blocking composition] \label{Def:Non-block}
%Let \(A_i = \OAg[i]\) with \( 0 \leq i \leq n\) be a family of open automata.
%Let \(A = A_0\mdbrk{j_i \mapsto A_i \middle| 1 \leq i \leq n}\) and $A= \OAg$.
%
%The composition \(A_0\mdbrk{j_i \mapsto A_i \middle| 1 \leq i \leq n}\) is non-blocking if   \(A \) has a reachability predicate such that, for each reachable configuration, if there is a possible transition in \(A_0\) then there is a possible transition in \(A\):
%\begin{multline*}
% \forall s = \mpar{s_0, s_1,..,s_n} \in S, V \uplus \biguplus_\subbox{t_0 \in \fOT{s_0}} \fvars{t_0} \vdash \\ \left(\reach{A}\mpar{s} \wedge \bigvee_\subbox{t_0 \in \fOT{s_0}} \fguard{t_0} \implies \bigvee_\subbox{t \in \fOT{s}} \fguard{t} \right)
%\end{multline*}
%\end{definition}
\begin{definition}[Non-blocking composition] \label{Def:Non-block}
Consider two OA:\\ \(A_1 \defobject \OAg[1]\) and \( A_2 \defobject \OAg[2]\).
Let $A$ be the OA resulting from the composition \(A = A_1\subst{A_2}{k}= \OAg\). 
The composition \(A_1\subst{A_2}{k}\) is non-blocking if   \(A \) has a reachability predicate such that, for each reachable configuration, if there is a possible transition in \(A_1\) then there is a possible transition in \(A\):
%\begin{multline*}
\[ \forall s = \mpar{s_1, s_2} \in S, V \uplus \biguplus_\subbox{t \in \fOT{s_1}} \fvars{t} \vdash  \left(\reach{A}\mpar{s} \wedge \bigvee_\subbox{t \in \fOT{s_1}} \fguard{t} \implies \bigvee_\subbox{t \in \fOT{s}} \fguard{t} \right)
\]
%\end{multline*}
\end{definition}
Like in the definition of simulation (Definition~\ref{def:HoleEqualSim}) we use guards to ensure that the transition can occur. 
In general, one would not want to only consider non-blocking composition as it may reveal a bit restrictive, but it is the best necessary condition that can be identified for compositionality of simulation. It will be used to prove composition theorems below. 
In absence of non-blocking composition, simulation may also be checked specifically for a given composed automaton.
%It is not sufficient to ensure the existence of equivalent transitions in general, but it will be sufficient in the context of refinement.

@@

We now  state the properties of our simulation, proofs of the properties can be found in the appendices. We express these properties in terms of hole-tracking simulation because, thanks to Property~\ref{lem-sim-sim}   all the properties  of hole-tracking simulation are also valid  for hole-equal simulation.
The first crucial property of  simulation  is that it is  a  preorder on the set of OA. This property enables stepwise refinement.

\begin{theorem}[Simulation is a preorder]
Hole-tracking simulation
   is reflexive  and  transitive:  it  is  a  preorder on the set of OA.
\end{theorem}

@@

\begin{example}  Consider again the \texttt{prod-cons} and \texttt{simprod-cons}  automata given in the examples above.  Since  \texttt{prod-cons} $\leq_{\{\texttt{P,Q}\}}$ \texttt{simprod-cons},  then according to Theorem \ref{thm:ContextRefinement}, 
\texttt{prod-cons[producer/P]} $\leq_{\{\texttt{Q}\}}$  \texttt{simprod-cons[producer/P]}. 
The automaton of \texttt{prod-cons[producer/P]} is shown in Fig.~\ref{Fig:ComposeOA}.
The automaton resulting  from the composition of  \texttt{Simprod} and \texttt{producer} is bigger and not shown here. \qed
\end{example}

@@

There are not a lot of works that study refinement for open systems. Defining refinement of open systems as trace inclusion  is  addressed  as a notion of subtyping in type theory 
(e.g. \cite{GayH:2005,BravettiZ:2021}). The definition of refinement is based on a connection  between session types and communicating automata theories -- a notion of session automata
based on Communicating Finite-State Machines, that are used for  modelling processes communicating through FIFO channels.    The refinement of open systems is also defined in terms of  alternating simulation \cite{Alur:1998,deAlfaro:2021}.
Alternating simulation  is originating from the game theory \cite{deAlfaro:2003}, it allows  the study of relation between individual components by viewing them as alternating transition systems. In particular,  a refinement of game-based automata expresses that the refined component can offer more services (input actions) and fewer service demands (output actions). However, the composition of such automata may
lead to illegal states, where one automaton issues an output that is not acceptable as input in the other one. The theory of alternating simulation provides an optimistic approach to compute compatibility between automata based on the fact that each automaton expects the other to provide  legal inputs, i.e, two components can be composed if there is an environment where they can work together. Our approach  has some commonalities with  the above mentioned simulation \cite{deAlfaro:2021}: both are process-oriented approaches even if they are not based on the same notion of simulation, and both include in the model  how to compose and interact with processes that are accepted as parameters.   Nevertheless, they differ in that our approach focuses on the compositional properties of the simulation, and {not on the fact that entities can be composed}.
% \LUDO{\st{TODO: what do we do better . in a different context?}}.

@@

\section{Proof of Transitivity for Refinement}\label{sec:proof-transitivity}
\begin{quote}
If $\wrel{A_1}{A_2}{H}$ and $\wrel{A_2}{A_3}{H'}$, then $\wrel{A_1}{A_3}{H\cap H'}$.
\end{quote}

@@

\item 
 Secondly, we need to prove that for any open transition $t_1$ in $T_1$  originating from $s_1$:
\begin{align*}
\OTx{1}{}{1}{1} \in \fOT{s_1}
\end{align*}
there exists an indexed family of OTs originating from $s_{3}$: 
\begin{multline*}
V_1 \uplus  V_3 \uplus \fvars{t_1}
\vdash\\\hspace{2em}
\mpar{R_{13}\mpar{s_1, s_3} \wedge g_1 \implies 
\operatorname*{\bigsymb{\bigvee}}_{z \in Z}
\mpar{\everymath{\displaystyle}
\begin{array}{l}
\alpha_1=\alpha_{3z}\wedge \bigwedge_\subbox{j \in J'_{3z} \cap (H\cap H')} \beta_{1j}=\beta_{3jz} \wedge \\g_{3z}\wedge 
{R_{12}\mpar{s'_1,s'_{3z}}}\psubst{\psi_1  \uplus \psi_{3z}}
\end{array}}}
\end{multline*}	
\RAB{The proof of this simulation step follows the same principles as the one given in \cite{AMHEEMA:2023} for the proof of transitivity of bisimulation\footnote{In particular, the bisimulation relation exhibited for proving transitivity is the same.}.  It relies on  the fact that one open transition of $A_1$ is simulated by a
family of open transitions of $A_2$, and one open transition of $A_2$ is simulated by a family of open transitions of $A_3$. Transitivity leads to a doubly indexed family of simulating open transition.}

@@

\section{Proof of Congruence}
\begin{quote}
Suppose $\wrel{A_2}{A_3}{H}$ and  $k \in H$.  and the composition $A_1\subst{A_2}{k}$ is non-blocking. We have: \[\wrel{A_1\subst{A_2}{k}}{A_1\subst{A_3}{k}}{J_1 \uplus H \setminus \mbrc{k}}\]
\end{quote}
\smallskip
Pictorially, the theorem states the following result:

@@

Let $R$ be the refinement relation relating states of $A_2$ and $A_3$. 
Let us denote with $t^2$ and $t^3$  the elements of $A_2$ and $A_3$ respectively.
Consider any two states $s_{12} = \mpar{s^1_1,s_2}$ and $s_{13} = \mpar{s^2_1,s_3}$ ($s^1_1$ and $s^2_1$  are the same with renaming). We define a relation $\mathcal{R}'$ relating states of $s_{12}$ and $s_{13}$ as follows:
\[\mathcal{R}' \mpar{s_{12}, s_{13}} = \mathcal{R}\mpar{s_{2}, s_{3}} \land \reach{A_{12}}(s_{12}) \wedge \bigwedge_\subbox{v_1\in V_1}
 v^1_1=v^2_1\wedge s^1_1 = s^2_1\]
Let us denote $H'=H\cup J_1 \setminus\{k\}$.
We want to prove that $\mpar{\mathcal{R}', H'}$  is a hole-tracking simulation of $A_{12}$ and $A_{13}$. As for the previous proof we deal with cases:
\begin{enumerate}
\item First, we have to prove the relation for initial states:
\[\sigma_{012} \uplus \sigma_{013} \vdash \mathcal{R}'\mpar{s_{012}, s_{013}}\]
with $\sigma_{012} = \sigma_{01}^1 \uplus \sigma_{03}$, $\sigma_{013} = \sigma_{01}^2 \uplus \sigma_{03}$, $s_{012}=(s_{01}^1,s_{02})$, and $s_{013}=(s_{01}^2,s_{03})$.\\
\RAB{The proof of this point is similar to that of Theorem \ref{thm:ContextRefinement}.}
\item Second, we need to prove for any OT $t_{12}$ in $T_{12}$ originating from $s_{12}$

>>>>

\keywords{Labelled transition systems   \and Simulation \and Refinement \and Composition.}
\end{abstract}
%
%
%
\section{Introduction}
%
%
%\TODO{TODO list}
%
%
%* \st{example a terminer}
%
%* choisir terminologie: refinement or simulation
%
%* \st{open automata  / Open Automata -> only use open automata or OA.}
%
%* positioning
%
%* \st{format according to conf: lncs FACS or SEFM}
%
%* global rewriting (partly done)

@@

%\begin{definition}[Expression algebra, Action algebra, Formulas, Terms]
Our expression algebra \(E\) is the disjoint union  of  terms,  actions, and  formulas
\( E=\terms \uplus \actions \uplus \formulas\).
\(\terms\) and \(\actions\) are term algebras.
The set of formulas \(\formulas\) contain at least first order formulas and equality\footnote{Equality does not need to be only syntactic.} over \(\terms\) and \(\actions\). 
%\end{definition}
For \(e \in E\),
 \(\fvars{e}\) is the set of variables in  $e$ that are not bound by a binder. 
An expression is closed if \(\fvars{e}=\emptyset\).
The set \(\values\) denotes values which is a subset of closed terms. \(\rformulas[V]\) is the set of formulas $f$ that only uses variables in $V$, i.e., the formulas such that  \(\fvars{f}\subseteq V\).
%The substitution in \(e \in E\) of \(x \in \fvars{e}\) by \(t \in \terms\), is denoted \(e\subst{t}{x}\), and its generalisation to t
The parallel substitution of variables in \(V\) by a map \(\psi: V \to \terms\) is denoted \(e\psubst{\psi}\).

@@

The OA  modelling the behaviour of such a system  using an  unbounded circular/ring buffer is depicted in Fig. \ref{Fig:RefineOA}.  The automaton has a single state with  two holes: $\texttt{P}$ and  $\texttt{Q}$ that are the two interacting processes.  $\texttt{l}$ (as last)  indicates the next available position for  enqueuing an element  and  $\texttt{f}$ (as first) is the position that contains the next element to be dequeued. %\footnote{The  arrow  without any state of origin is a graphical convention to indicate that the pointers are initialized at the beginning}. 
The buffer reacts to a push from $\texttt{P}$ and enqueues it. Similarly, whenever   $\texttt{Q}$ pops an element, it dequeues it. Additionally, whenever $\texttt{Q}$ produces an item, it is exposed as an external observable  $\texttt{print}$ action. 
When any process do its internal computation, it is exposed externally as unobservable action $\tau$.

@@

\begin{figure}[!tb]
 \centering
   \includegraphics[width=.99\textwidth]{Figures/FIFORing.pdf}
   \caption{OA for the \texttt{prod-cons} system using FIFO circular buffer.
    \label{Fig:RefineOA}} 
\end{figure}
\qed
\end{example}

@@

\begin{example} \label{Example:prodandcons} Fig. \ref{Fig:procandcons} shows a producer automaton  and a consumer  automaton that can be used  to fill the holes $\texttt{P}$ and $\texttt{Q}$ of \texttt{prod-cons} defined in Example~\ref{Example:fifo-system}.
 \begin{figure}[!t]
 \centering
   \includegraphics[width=.35\textwidth]{Figures/P-proc.pdf}\hfill 
   \includegraphics[width=.45\textwidth]{Figures/Q-proc.pdf}
   \caption{(Left) A \texttt{producer}.  It produces one item at a time and pushes it.  (Right)  A \texttt{consumer}. It pops an item, does some work and pushes the result. \label{Fig:procandcons}}
\end{figure}

@@

\begin{figure}[!t]
 \centering
   \includegraphics[width=.99\textwidth]{Figures/compos-FIFO-Pro.pdf}
   \caption{OA for \texttt{prod-cons}[\texttt{P}/\texttt{producer}]. 
   \label{Fig:ComposeOA}} 
\end{figure}
\end{example}

\subsection{Relations between Open Automata}
Establishing semantic equivalences and simulation relations between different OA requires  to compare their states. For this purpose, we suppose that the variables of the two OA are disjoint (a renaming of variables may have to be applied before comparing OA states).
\begin{definition}[Relation on open automata configurations] Suppose $V_1$ and $V_2$ are disjoint.
A relation on configurations of two OA \(\OAg[1]\) and \(\OAg[2]\) is a function \( \mathcal{R}: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\).
\end{definition}
%\TODO{C'est etrange, la formulation "the variables they refer to", comme il n'y a plus de variables d'etat... +1}
The idea is that two states are related depending on the satisfiability of the expression relying their variables, i.e., if the variables of the OA verify a certain formula. 
In other words, to each pair of states is attached a boolean formula that may refer to the  variables of each of the two OA, stating whether the two states are related or not.
Additionally, we say that the relation $ \mathcal{R}$ \emph{relates the initial states} of the automata if: \(\sigma_{01} \uplus \sigma_{02} \vdash  \mathcal{R}\mpar{s_{01}, s_{02}}\).

@@

Reachability takes into account all paths, and can over-approximate the reachable configurations. 
From an automation point of view, finding the most precise reachability predicate for a given automaton is not decidable because of the symbolic nature of OA, but only an over-approximation is necessary.

@@

Similarly to FH-bisimulation~\cite{AMHEEMA:2023} we are interested  in finding simulation relations between configurations of two OA that contain variables and holes. When dealing with open systems it is common to define simulation in terms of a simulation relation.
We rely on a classical notion of simulation inspired from~\cite{AMHEEMA:2023}. The idea is to consider two configurations related by a relation; if one state can do a transition, then the other can also make this transition. 
Like for bisimulation, a simulation relation characterises when two states are related, and this  characterisation is expressed as a predicate on the variables of the two automata.
Simulation defines conditions on a relation $\mathcal{R}$ such that $\mathcal{R}\mpar{s_1, s_2}$ is a predicate (possibly involving variables of the automata) that is true when the state $s_1$ of $A_1$ simulates the state $s_2$ of $A_2$.

@@

\begin{example} \label{example:fifo-system-one}
To illustrate the simulation  of OA, we consider a variation on the \texttt{prod-cons} example. Namely,  we suppose that  the two processes  \texttt{P} and  \texttt{Q} communicate through a one-place  buffer. Fig. \ref{Fig:SpecOA} shows the OA modelling this simpler version of the system, that we refer to as \texttt{simprod-cons}. 
\begin{figure}[!b]
 \centering
   \includegraphics[width=.90\textwidth]{Figures/FIFOpen2.pdf}
   \caption{  The \texttt{simprod-cons} OA: the system using one-place  buffer.  \label{Fig:SpecOA}}
\end{figure}
We can easily check that this automaton simulates the one of Fig.~\ref{Fig:RefineOA}.
Indeed, one can see that  $\mathcal{R}=\{(r_0,s_0) \mapsto \texttt{l}=\texttt{f}, (r_0,s_1) \mapsto \texttt{f}=\texttt{l}+1\% \texttt{N} \}$ is  a  simulation 
relation. It follows that  \texttt{prod-cons} $\leq$ \texttt{simprod-cons}. \qed
\end{example}

@@

\medskip
One major challenge in the extension of simulation to different sets of holes is to maintain a form of transitivity while being able to take into account the actions of some of the holes. A naive definition of simulation would ensure that only the holes that are identical in the two OA are taken into account in the simulation. Unfortunately, considering all the common holes does not ensure transitivity of the simulation for the following reason. If $A_1$ simulates $A_2$ and $A_2$ simulates $A_3$, and one hole $j$ appears in $A_3$ and in $A_1$ but not in $A_2$ then we have no guarantee on the way $A_1$ and $A_3$ take the actions of this his hole into account, thus  a simulation between and $A_1$ and $A_3$ would require conditions involving actions of the hole $j$ which cannot be ensured. The way we solve this issue is to remember in the simulation relation which holes have been compared. This makes the relation parameterised by a subset of the set of holes that belong to the two automata that we want to take into account.
This way, in the example above, we would have no guarantee on actions the hole $j$ by transitivity but can state a simulation relation with guarantees on the actions of the other holes.

@@

\begin{property}[Relating simulations]\label{lem-sim-sim}
Hole-equal simulation is a particular case of hole-tracking simulation when  $J_1=J_2=H$.
\end{property}
\begin{example} Consider the automata of Examples \ref{Example:fifo-system} and \ref{example:fifo-system-one}. As we saw above, \texttt{prod-cons} $\leq$ \texttt{simprod-cons}, therefore \texttt{prod-cons} $\leq_{\{\texttt{P,Q}\}}$ \texttt{simprod-cons}.
\end{example}
%\TODO{potential ex 4: check if the example with a hole filled is a simulation of the example with holes? }

@@

Before reasoning on the properties of simulation. we need to introduce one additional notion that characterises when the composition of two automata does not introduce new blocked transitions.

@@

%\begin{definition}[Non-blocking composition] \label{Def:Non-block}
%Let \(A_i = \OAg[i]\) with \( 0 \leq i \leq n\) be a family of open automata.
%Let \(A = A_0\mdbrk{j_i \mapsto A_i \middle| 1 \leq i \leq n}\) and $A= \OAg$.
%
%The composition \(A_0\mdbrk{j_i \mapsto A_i \middle| 1 \leq i \leq n}\) is non-blocking if   \(A \) has a reachability predicate such that, for each reachable configuration, if there is a possible transition in \(A_0\) then there is a possible transition in \(A\):
%\begin{multline*}
% \forall s = \mpar{s_0, s_1,..,s_n} \in S, V \uplus \biguplus_\subbox{t_0 \in \fOT{s_0}} \fvars{t_0} \vdash \\ \left(\reach{A}\mpar{s} \wedge \bigvee_\subbox{t_0 \in \fOT{s_0}} \fguard{t_0} \implies \bigvee_\subbox{t \in \fOT{s}} \fguard{t} \right)
%\end{multline*}
%\end{definition}
\begin{definition}[Non-blocking composition] \label{Def:Non-block}
Consider two OA:\\ \(A_1 \defobject \OAg[1]\) and \( A_2 \defobject \OAg[2]\).
Let $A$ be the OA resulting from the composition \(A = A_1\subst{A_2}{k}= \OAg\). 
The composition \(A_1\subst{A_2}{k}\) is non-blocking if   \(A \) has a reachability predicate such that, for each reachable configuration, if there is a possible transition in \(A_1\) then there is a possible transition in \(A\):
%\begin{multline*}
\[ \forall s = \mpar{s_1, s_2} \in S, V \uplus \biguplus_\subbox{t \in \fOT{s_1}} \fvars{t} \vdash  \left(\reach{A}\mpar{s} \wedge \bigvee_\subbox{t \in \fOT{s_1}} \fguard{t} \implies \bigvee_\subbox{t \in \fOT{s}} \fguard{t} \right)
\]
%\end{multline*}
\end{definition}
Like in the definition of simulation (Definition~\ref{def:HoleEqualSim}) we use guards to ensure that the transition can occur. 
In general, one would not want to only consider non-blocking composition as it may reveal a bit restrictive, but it is the best necessary condition that can be identified for compositionality of simulation. It will be used to prove composition theorems given below. 
In absence of non-blocking composition, simulation may also be checked specifically for a given composed automaton.
%It is not sufficient to ensure the existence of equivalent transitions in general, but it will be sufficient in the context of refinement.

@@

We now  state the properties of our simulation, their proofs  can be found in the appendices. We express these properties  in terms of hole-tracking simulation because, thanks to Property~\ref{lem-sim-sim}   all the properties  of hole-tracking simulation are also valid  for hole-equal simulation.
The first crucial theorem of  simulation  is that it is  a  preorder on the set of OA. This latter enables stepwise refinement.

\begin{theorem}[Simulation is a preorder]\label{thm:preoder}
Hole-tracking simulation
   is reflexive  and  transitive:  it  is  a  preorder on the set of OA.
\end{theorem}

@@

\begin{example}  Consider again the \texttt{prod-cons} and \texttt{simprod-cons}  automata given in the examples above.  Since  \texttt{prod-cons} $\leq_{\{\texttt{P,Q}\}}$ \texttt{simprod-cons},  then according to Theorem \ref{thm:ContextRefinement}, 
\texttt{prod-cons[producer/P]} $\leq_{\{\texttt{Q}\}}$  \texttt{simprod-cons[producer/P]}. 
The automaton of \texttt{prod-cons[producer/P]} is shown in Fig.~\ref{Fig:ComposeOA}. While the automaton resulting  from the composition of  \texttt{simprod-cons} and \texttt{producer} is bigger and not shown here. \qed
\end{example}

@@

There are not a lot of works that study refinement for open systems. Defining refinement of open systems as trace inclusion  is  addressed  as a notion of subtyping in type theory 
(e.g. \cite{GayH:2005,BravettiZ:2021}). The definition of refinement is based on a connection  between session types and communicating automata theories -- a notion of session automata
based on Communicating Finite-State Machines, that are used for  modelling processes communicating through FIFO channels.    The refinement of open systems is also defined in terms of  alternating simulation \cite{Alur:1998,deAlfaro:2021}.
Alternating simulation  is originating from the game theory \cite{deAlfaro:2003}, it allows  the study of relation between individual components by viewing them as alternating transition systems. In particular,  a refinement of game-based automata expresses that the refined component can offer more services (input actions) and fewer service demands (output actions). However, the composition of such automata may
lead to illegal states, where one automaton issues an output that is not acceptable as input in the other one. The theory of alternating simulation provides an optimistic approach to compute compatibility between automata based on the fact that each automaton expects the other to provide  legal inputs, i.e., two components can be composed if there is an environment where they can work together. Our approach  has some commonalities with  the above mentioned simulation \cite{deAlfaro:2021}: both are process-oriented approaches even if they are not based on the same notion of simulation, and both include in the model  how to compose and interact with processes that are accepted as parameters.   Nevertheless, they differ in that our approach focuses on the compositional properties of the simulation, and {not on the fact that entities can be composed}.
% \LUDO{\st{TODO: what do we do better . in a different context?}}.

@@

\section{Proof of Transitivity for Refinement (Theorem~\ref{thm:preoder})}\label{sec:proof-transitivity}
\begin{quote}
If $\wrel{A_1}{A_2}{H}$ and $\wrel{A_2}{A_3}{H'}$, then $\wrel{A_1}{A_3}{H\cap H'}$.
\end{quote}

@@

\item 
 Secondly, we need to prove that for any open transition $t_1$ in $T_1$  originating from $s_1$:
\begin{align*}
\OTx{1}{}{1}{1} \in \fOT{s_1}
\end{align*}
there exists an indexed family of OTs originating from $s_{3}$: 
\begin{multline*}
V_1 \uplus  V_3 \uplus \fvars{t_1}
\vdash\\\hspace{2em}
\mpar{R_{13}\mpar{s_1, s_3} \wedge g_1 \implies 
\operatorname*{\bigsymb{\bigvee}}_{z \in Z}
\mpar{\everymath{\displaystyle}
\begin{array}{l}
\alpha_1=\alpha_{3z}\wedge \bigwedge_\subbox{j \in J'_{3z} \cap (H\cap H')} \beta_{1j}=\beta_{3jz} \wedge \\g_{3z}\wedge 
{R_{12}\mpar{s'_1,s'_{3z}}}\psubst{\psi_1  \uplus \psi_{3z}}
\end{array}}}
\end{multline*}	
The proof of this simulation step follows the same principles as the one given in \cite{AMHEEMA:2023} for the proof of transitivity of bisimulation\footnote{In particular, the bisimulation relation exhibited for proving transitivity is the same.}.  It relies on  the fact that one open transition of $A_1$ is simulated by a
family of open transitions of $A_2$, and one open transition of $A_2$ is simulated by a family of open transitions of $A_3$. Transitivity leads to a doubly indexed family of simulating open transition.

@@

\section{Proof of Congruence (Theorem~\ref{thm:CongurenceRefinement})}
\begin{quote}
Suppose $\wrel{A_2}{A_3}{H}$ and  $k \in H$.  and the composition $A_1\subst{A_2}{k}$ is non-blocking. We have: \[\wrel{A_1\subst{A_2}{k}}{A_1\subst{A_3}{k}}{J_1 \uplus H \setminus \mbrc{k}}\]
\end{quote}
\smallskip
Pictorially, the theorem states the following result:

@@

Let $R$ be the refinement relation relating states of $A_2$ and $A_3$. 
Let us denote with $t^2$ and $t^3$  the elements of $A_2$ and $A_3$ respectively.
Consider any two states $s_{12} = \mpar{s^1_1,s_2}$ and $s_{13} = \mpar{s^2_1,s_3}$ ($s^1_1$ and $s^2_1$  are the same with renaming). We define a relation $\mathcal{R}'$ relating states of $s_{12}$ and $s_{13}$ as follows:
\[\mathcal{R}' \mpar{s_{12}, s_{13}} = \mathcal{R}\mpar{s_{2}, s_{3}} \land \reach{A_{12}}(s_{12}) \wedge \bigwedge_\subbox{v_1\in V_1}
 v^1_1=v^2_1\wedge s^1_1 = s^2_1\]
Let us denote $H'=H\cup J_1 \setminus\{k\}$.
We want to prove that $\mpar{\mathcal{R}', H'}$  is a hole-tracking simulation of $A_{12}$ and $A_{13}$. As for the previous proof we deal with cases:
\begin{enumerate}
\item First, we have to prove the relation for initial states:
\[\sigma_{012} \uplus \sigma_{013} \vdash \mathcal{R}'\mpar{s_{012}, s_{013}}\]
with $\sigma_{012} = \sigma_{01}^1 \uplus \sigma_{03}$, $\sigma_{013} = \sigma_{01}^2 \uplus \sigma_{03}$, $s_{012}=(s_{01}^1,s_{02})$, and $s_{013}=(s_{01}^2,s_{03})$.\\
The proof of this point is similar to that of Theorem \ref{thm:ContextRefinement}.
\item Second, we need to prove for any OT $t_{12}$ in $T_{12}$ originating from $s_{12}$
